---
title: "brenz24-03-growth_analysis"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output: github_document
params:
  name: "brenz24-03-growth_analysis" # change if you rename file
---

```{r here, message=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "3d5f015f-5bf7-42d2-991b-3568bbd18592")
```

The purpose of this document is ...

```{r packages}
library("tidyverse")
library("DGrowthR")
```

```{r directories}
# create or *empty* the target directory, used to write this file's data: 
projthis::proj_create_dir_target(params$name, clean = TRUE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```

## Read pre-processed DGrowthR objects

```{r read.objects}
dg.logged <- readRDS(path_source("brenz24-01-preprocess_data", "dgobj_logged.rds"))

```

## Low-dimensional representation. 
 
```{r estimate.lowdim}

# FPCA
dg.logged <- estimate_fpca(dg.logged)

# UMAP
set.seed(123)
dg.logged <- estimate_umap(dg.logged)

```


```{r plot.fpca}

plot_fpca(dg.logged) +
  labs(title = "Logged")

``` 
  

```{r plot.umap}

plot_umap(dg.logged) +
  labs(title = "Logged")

```

## Clustering. 
  
```{r clustering}

dg.logged <- clustering(dg.logged, k=0.015)

```
  
  
```{r logged.cluster.viz}
plot_fpca(dg.logged, color="cluster_assignment") +
  labs(title = "Logged")

plot_umap(dg.logged, color="cluster_assignment") +
  labs(title = "Logged")

plot_growth_curves(dg.logged, color = "cluster_assignment", facet="cluster_assignment")

```


## Growth parameters. 

Estimate growth parameters for each genotype-compound-concentration combination


  
```{r estimate.growth.parms}


dg.logged.gparams <- estimate_growth_parameters(dg.logged, 
                                        model_covariate = "genotype_well",
                                        save_gp_data = TRUE,
                                        n_cores=4)

```
  



```{r cluster.assignment.mv}
treatment.cluster <- dg.logged@metadata %>% 
  left_join(dg.logged@cluster_assignment, by="curve_id") %>% 
  
  group_by(genotype_well) %>% 
  count(cluster) %>% 
  filter(n == max(n)) %>% 
  ungroup() %>% 
  
  distinct(genotype_well, .keep_all = TRUE) %>% 
  select(genotype_well, cluster)
```



```{r plot.gparams}

dg.logged.gparams@growth_parameters %>% 
  rename("genotype_well" = "gpfit_id") %>% 
  
  left_join(treatment.cluster, by="genotype_well") %>% 
  filter(!is.na(AUC) & !is.na(max_growth_rate)) %>% 
  mutate(normalized_max_growth_rate = (max_growth_rate - min(max_growth_rate)) / (max(max_growth_rate) - min(max_growth_rate)),
         normalized_AUC = (AUC - min(AUC)) / (max(AUC) - min(AUC))) %>% 
  
  
  ggplot(aes(x=normalized_AUC, y=normalized_max_growth_rate, color=cluster)) +
  geom_point() +
  theme_bw()

```

  
## Multiple comparisons dCBASS vs WT


```{r drug.metadata}
drug.metadata <- dg.logged.gparams@metadata %>% 
  select(well, Drug, ConcMock) %>% 
  distinct() %>% 
  
  unite("drug_conc", c(Drug, ConcMock), sep="-")

```


 
Multiple growth testing. 

```{r read.growth.comparisons}

# Results
dg.results.p1 <- readRDS("permutation_results/brenzinger_500p.rds")
#dg.results.p2 <- readRDS("tmp_permutations/brenzinger_1000p.rds")


# Read first 500 permutation
dg.permutations.p1 <- readRDS("permutation_results/brenzinger_dgresults.rds")

# Read second 500 permutations
dg.permutations.p2 <- readRDS("permutation_results/brenzinger_dgresults_1000p.rds")

# Join permutations
dg.permutations <- dg.permutations.p1 %>% 
  left_join(dg.permutations.p2, by="comparison")

# Summary stats for permutations
perm.sumstats.p1 <- dg.permutations.p1 %>% 
  pivot_longer(cols = -comparison, names_to = "nPerm", values_to = "BF_value") %>% 
  
  group_by(comparison) %>%
  summarise(median_BF = median(BF_value),
            mean_BF = mean(BF_value))

dg.results.p1 <- dg.results.p1 %>% 
  left_join(perm.sumstats.p1, by="comparison")
```


```{r gamma.approximations}
# Make sure everything is in the same order
dg.joined <- dg.permutations %>% 
  left_join(dg.results.p1 %>% select(comparison, likelihood_ratio), by="comparison") 

# Calculate the gamma approximation
BF_obs <- dg.joined$likelihood_ratio
BF_perm <- dg.joined %>% 
  select(-c(comparison, likelihood_ratio)) %>%
  as.matrix()

papprox_gamma <- permAprox::permaprox(
    alternative = "twoSided",
    nullValue = "median",
    tPerm = BF_perm,
    tObs = BF_obs,
    method = "gamma",
    fitThresh = 0.2,
    gofTestGamma = FALSE,
    includeObs = FALSE,
    multAdj = "BH",
    cores = 6
  )

dg.permutations <- dg.permutations %>% 
  mutate("gamma_pval" = papprox_gamma$p,
         
         "emp_pval" = papprox_gamma$pEmp)
```




```{r}
dg.results.gpval <- dg.results.p1 %>%
  select(comparison, AUC.FoldChange, likelihood_ratio, median_BF, mean_BF) %>% 
  
  left_join(dg.permutations %>% select(comparison, emp_pval, gamma_pval), by="comparison") %>% 
  filter(likelihood_ratio > median_BF) %>%
  mutate(logAUC.FC = log2(AUC.FoldChange),
         gamma_adj_pval = p.adjust(gamma_pval, method="BH"),
         emp_adj_pval = p.adjust(emp_pval, method="BH"))

dg.results.gpval %>% 
  ggplot(aes(x=gamma_pval)) +
  geom_histogram(binwidth = 0.01)
```

```{r}
diff.growers <- dg.results.gpval %>% 
  filter(gamma_adj_pval <= 0.1,
         logAUC.FC > 0 | logAUC.FC <= -0.25) %>% 
  
  mutate(well = str_extract(comparison, "genotype_well: dCBASS_(.*?) v.s. wt_.*?", group = 1)) %>% 
  left_join(drug.metadata, by="well") %>% 
  
  select(comparison, drug_conc, logAUC.FC, emp_pval, emp_adj_pval, gamma_pval, gamma_adj_pval) %>% 
  arrange(desc(logAUC.FC))
  
diff.growers
```


```{r volcano.plot}
ggplot(dg.results.gpval, aes(y=-log10(gamma_adj_pval), x=logAUC.FC, fill=logAUC.FC)) + 
  geom_point(alpha=0.85, shape=21, color="black") +  # The alpha parameter controls the transparency of the points
  
  
  scale_fill_gradient2(low = "blue", # Here we tell ggplot to use a diverging color scale",
                        mid = "#f7f7f7",
                        high = "red") +
  
  #geom_text_repel(data = diff.growers, aes(label=drug_conc), 
  #                max.overlaps = Inf, size=2.75, min.segment.length = 0, color="black", 
  #                box.padding = 0.7, point.padding = 0.2, fontface="plain") +
  
  geom_vline(xintercept = 0, color="black", linetype="longdash") +
  geom_vline(xintercept = 0.25, color="black", linetype="longdash") +
  geom_vline(xintercept = -0.25, color="black", linetype="longdash") +
  
  geom_hline(yintercept = -log10(0.05), color="black", linetype="longdash") +
  theme_bw()
  
```



## Comparing the effect of dCBASS to wild-type




Comparing the AUC. 
```{r}
library(ggrepel)
```


```{r compare.auc}
# Gather relevant data
auc.comparison.df <- dg.logged.gparams@growth_parameters %>% 
  select(gpfit_id, AUC) %>% 
  separate(gpfit_id, into=c("genotype", "well")) %>% 
  
  pivot_wider(id_cols = well, names_from = genotype, values_from = AUC) %>% 
  mutate(auc.diff = dCBASS - wt) %>% 
  
  left_join(drug.metadata, by="well")



# Gather top hits
top.auc.diff <- auc.comparison.df %>% 
  slice_max(order_by = auc.diff, n=4)


# Plot
ggplot(auc.comparison.df, aes(x=wt, y=dCBASS, fill=auc.diff)) +
  
  geom_abline() +
  geom_point(shape=21, color="black") +
  
   scale_fill_gradient2(low = "#1065ab",
                        mid = "#f9f9f9",
                        high = "#b31529",
                        midpoint = 0) +
  
  
  geom_text_repel(data = top.auc.diff, aes(label=drug_conc), 
                  max.overlaps = Inf, size=3, min.segment.length = 0, color="black", box.padding = 0.7, point.padding = 0.2, fontface="plain") +
  
  theme_bw() +
  labs(x="Wild-type AUC",
       y=latex2exp::TeX("$\\Delta$CBASS\\ AUC"),
       fill="Diff. AUC")



```


```{r growth.comparison.amoxi}

gc.amoxi <- growth_comparison(dg.logged, 
                              comparison_info = c("genotype_well", "dCBASS_A11" ,"wt_A11"),
                              save_gp_data = TRUE,
                              permutation_test = FALSE)


gc.amoxi.plots <- plot_growth_comparison(gc.amoxi)
gc.amoxi.plots$alternative
gc.amoxi@growth_comparison$result
```
  


```{r}
gc.sulfa <- growth_comparison(dg.logged, 
                              comparison_info = c("genotype_well", "dCBASS_O1" ,"wt_O1"),
                              save_gp_data = TRUE,
                              permutation_test = FALSE)


gc.sulfa.plots <- plot_growth_comparison(gc.sulfa)
gc.sulfa@growth_comparison$result
gc.sulfa.plots$alternative
```


```{r compare.deathrate}
# Gather relevant data
dr.comparison.df <- dg.logged.gparams@growth_parameters %>% 
  
  mutate(max_death_rate = if_else(is.na(max_death_rate), 0, max_death_rate)) %>% 
  select(gpfit_id, max_death_rate) %>% 
  separate(gpfit_id, into=c("genotype", "well")) %>% 
  
  pivot_wider(id_cols = well, names_from = genotype, values_from = max_death_rate) %>% 
  mutate(dr.diff = dCBASS - wt) %>% 
  
  left_join(drug.metadata, by="well")



# Gather top hits
top.dr.diff <- dr.comparison.df %>% 
  slice_max(order_by = abs(dr.diff), n=3)


# Plot
ggplot(dr.comparison.df, aes(x=wt, y=dCBASS, fill=dr.diff)) +
  
  geom_abline() +
  geom_point(shape=21, color="black") +
  
   scale_fill_gradient2(low = "#1065ab",
                        mid = "#f9f9f9",
                        high = "#b31529",
                        midpoint = 0) +
  
  
  geom_text_repel(data = top.dr.diff, aes(label=drug_conc), 
                  max.overlaps = Inf, size=3, min.segment.length = 0, color="black", box.padding = 0.7, point.padding = 0.2, fontface="plain") +
  
  theme_bw() +
  labs(x="Wild-type Death rate",
       y=latex2exp::TeX("$\\Delta$CBASS\\ Death\\ rate"),
       fill="Diff. Death rate")



```

## Files written

These files have been written to the target directory, ```r paste0("data/", params$name)```:

```{r list-files-target}
projthis::proj_dir_info(path_target())
```
