---
title: "Get started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DGrowthR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---  

```{r, include = FALSE}
# Suppress title check warning
options(rmarkdown.html_vignette.check_title = FALSE)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.height=5, 
  fig.width=7)

```
    
This page aims to give an overview of `DGrowthR` and introduce you to the main functions that you might want to use for your own analysis. Broadly speaking, this tutorial is divided into the following sections:  
  
1. [Preparing your data for `DGrowthR`](#preparing-your-data-for-dgrowthr).   [Required]
2. Preparing a DGrowthR object. [Required]
3. Pre-processing. [Required]
4. Exploratory data analysis.  
5. Growth parameter calculation.  
6. Differential growth testing. 
7. Multiple differential growth testing.

```{r setup, message=FALSE}
library("tidyverse")

library("DGrowthR")
```

  
## Preparing your data for `DGrowthR`. 
  
To start our analysis we first need to prepare a `DGrowthR::DGrowthR()` object. To do this, we require two sources of information:  
1. Growth data. 
2. Metadata.

Below we will go into more details of each one.  
  
### Growth data. 
  
`DGrowthR` can be used to analyse growth data from multi-well plate readers. Depending on the specific platform that is used, the actual output of each machine might look slightly different. **`DGrowthR` assumes your data consists of simultaneous measurements of growth**, preferably that are regularly interspaced. This means that the growth across all wells is measured at the same time and, ideally, that the time between measurements is always the same. If the data is not measured at regular intervals, `DGrowthR` will still be able to analyze your data. Examples of the kinds of data the can be used can be seen in our [Source Data](https://github.com/bio-datascience/DGrowthR/tree/main/Source_Data) folder.   
  
** Below is an example of the data collected for one 384 well plate from [Brenzinger 2024](https://www.nature.com/articles/s41564-023-01556-y). This dataframe could be used directly to build a `DGrowthR::DGrowthR()` object. The important features include: **
  
1. A column called **Time**: The values in this column indicate the time at which the OD measurements were made for all of the wells in this plate. The units of the **Time** column (seconds, minutes, hours) are taken _as is_. No unit conversion is performed by `DGrowthR`.   
2. Columns indicating wells: All of the columns besides the **Time** column indicate the wells in the plate. The values are the measured optical density in each well at a specific time point.   
3. An arbitrary number of wells can be present (typically 384 or 96). Typically, each well is identified as a combination of an uppercase letter followed by a number, with no spaces. This information is used to identify each individual growth curve.  
  
[Source_Data/Brenzinger_2024/wt_1_OD.txt](https://github.com/bio-datascience/DGrowthR/blob/main/Source_Data/Brenzinger_2024/od_data/wt_1_OD.txt)
  
  
```{r source_example_brenzinger}
  
# Read example data from Brenzinger 2024
example_brenzinger <- read.delim("../Source_Data/Brenzinger_2024/od_data/wt_1_OD.txt", 
                                 sep = '\t')

tibble(example_brenzinger) %>% 
  rmarkdown::paged_table()

```
  
  
As stated above, this data frame could be used directly by `DGrowthR`.  
  
#### What if I want to analyze data from multiple plates?. 
  
This is definitely possible with `DGrowthR`. There are just a couple of things to keep in mind:  
  
1. The data for each plate should be in separate files.

2. It is assumed that measurements are taken at roughly the same timepoints across all plates. This means that the **Time** column should be roughly the same for all plates. Due to small variations between different plate measurements, the values in the **Time** column might be slightly different.   

2. Each growth curve will now be identified as the combination of {plate name}_{well}. 
  
In order to read the data from multiple plates, harmonize the **Time** column, and assign individual curve identifiers in a way that can be used by `DGrowthR`, we provide a standalone function called `read_multiple_plates()`. Below, we use it to read all of the data from the [Brenzinger 2024](https://www.nature.com/articles/s41564-023-01556-y) study. In this case we know that 20 measurements are taken; the time between the first and second measurement is ~2 hours, followed by regular 44 minute intervals. This can help us create a common time vector for all files.

```{r read.multiple.brenzinger2024}

# Create common time vector
common_time_vector <- c(0, 2, seq(from=2.739, by=0.739, length.out=18))
 
# Read all OD data
brenzinger_data <- read_multiple_plates(path_target = "../Source_Data/Brenzinger_2024/od_data/",
                                        time_vector = common_time_vector)

```
  
The output of this function is a list with two data frames. First is the "od_data" data frame. As you can see, it has a common **Time** column with each individual growth curve in remaining columns. Each growth is now identified with a **curve_id** which is built as {plate_name}_{well}, with the plate_name actually just being the name of the file from which the data was read, without the extension. In total there are 20 measurements for 2,304 (384 * 6) growth curves!
  
```{r complete.od.brenzinger}

# Show the first five columns of the complete OD data
brenzinger_od <- brenzinger_data[["od_data"]] 


brenzinger_od %>% 
  rmarkdown::paged_table()
  

```

Additionally, a "metadata" data frame is provided. This data frame explains where each **curve_id** is coming from. The plate name (file name without the extension) and the original well are explicitly mentioned.  
 
```{r brenzinger.metadata}

brenzinger_metadata <- brenzinger_data[["metadata"]] 


brenzinger_metadata %>% 
  rmarkdown::paged_table()

```

This sets us up nicely to now discuss the metadata.  
  
### Metadata. 
  
The metadata contains all of the information relevant to the experimental setup. The information contained here can be quite arbitrary and suit the needs of the researcher. For example, it can contain information about the conditions under which each growth curve was measured, replicate number, batch number, etc. Importantly, there must be a column named **curve_id**, that contains the names of all of the growth curves present in the data frame with the optical density information. 
  
In our case, we will use the information about genotype, compound and concentration provided by the original study. 

```{r brenzinger.metadata.info}

# Information about each plate
database.df <- read_tsv("../Source_Data/Brenzinger_2024/Database.txt", 
                        show_col_types = FALSE) %>% 
  separate(Strain, into = c("genotype", NA)) %>% 
  select(file_id, genotype, Rep_Nr) %>% 
  mutate(plate_name = tools::file_path_sans_ext(file_id))

# Information about each well
mapper.df <- read_tsv("../Source_Data/Brenzinger_2024/Map.txt",
                       show_col_types = FALSE) %>% 
  rename("well" = "Well")


# Add this information to the metadata data frame
brenzinger_metadata <- brenzinger_metadata %>% 
  left_join(database.df, by="plate_name") %>% 
  left_join(mapper.df, by="well") %>% 
  unite("genotype_well", c(genotype, well), sep="_", remove=FALSE)

brenzinger_metadata %>% 
  rmarkdown::paged_table()


```

  
Now we have a lot more information about the conditions under which each growth curve was gathered. With these sources of information, we can now build our `DGrowthR::DGrowthR()` object.   
  
## Building a DGrowth object. 
  
Once we are satisfied with our growth data and metadata data frames, we can instantiate a `DgrowthR()` object with the `DgrowthRFromData()` function.
  
```{r instantiate.dgobj}

# Instantiate a DGrowthR object
dgobj <- DGrowthRFromData(od_data = brenzinger_od,
                          metadata = brenzinger_metadata)
```
We now have a DGrowthR object. This is the heart of our package. All functions require a `DGrowthR` object as input. For example, we can quickly visually inspect our data using the `plot_growth_curves()` function.  
 
 
```{r plot.all.raw.curves}

plot_growth_curves(dgobj)

```
  
We can inspect our growth curves using variables of the metadata with the `facet` and `color` parameters. 
  
```{r plot.raw.facets}

plot_growth_curves(dgobj, facet="plate_name", color="genotype")

```
  
It's worth mentioning that all of the plots returned by `DGrowthR` functions are [ggplot](https://ggplot2.tidyverse.org/) objects. Meaning you may further customize them to your liking using ggplot syntax and functions.  
  
Now that our data is ready we can start pre-processing.  
  
## Pre-processing. 
  
Pre-processing growth curves is essential before performing any analysis. This is done by the `preprocess_data()` function. The following transformations are performed:  
  
1. First the user may specify if some of the initial growth measurements should be ignored with the `skip_first_n_timepoints` parameter. All growth curves would be shortened so they start at timepoint skip_first_n_timepoints + 1. 

2. The optical density measurements are log-transformed. This enables accurate estimation of growth parameters such as maximum specific growth rate, end of lag phase, etc. Both the log and linear scaled data are conserved.

3. Baseline correction is performed by subtracting the initial OD measurement from the rest of the growth curve. 
  
```{r preprocess.brenzinger}

dgobj <- preprocess_data(dgobj)

```

We can see how our data has been transformed. 
  
```{r plot.processed}

plot_growth_curves(dgobj, facet="plate_name", color="genotype")

```


Notice how now the y-axis starts at 0.  

  
## Low dimensional representations. 
  
Now that we have pre-processed data we can compute a low-dimensional representation of our data. `DGrowthR` offers the possibility to do this either with a functional PCA, or a UMAP embedding.
  
```{r estimate.lowdim.viz}

dgobj <- estimate_fpca(dgobj)
dgobj <- estimate_umap(dgobj)
```
  
Now we can plot both representations and see what we observe    

```{r plot.umaps.fpca}

plot_fpca(dgobj)


plot_umap(dgobj)

```

```{r}
plot_fpca(dgobj)


plot_umap(dgobj)
```


This is looking interesting. One can further investigate what is being captured by coloring a given covariate in the metadata.  
 
```{r plot.umaps.genotype}

plot_umap(dgobj, color = "genotype")

```

  
One can also visualize certain growth parameters that are calculated by DGrowthR. That is discussed in the next section.  
  
## Clustering. 
  
One can cluster growth curves based on their low-dimensinal embeddings in the fPCA or UMAP. This is acheived by the `
  
```{r cluster.gcs}

# Clustering based on FPCA coordinates
dgobj <- clustering(dgobj, embedding="fpca")


# Now you can plot cluster assignment
plot_fpca(dgobj, color = "cluster_assignment")
```

  
Since we know that we are clustering the individual growth curves, we can plot our growth curves with updated metadata

```{r updated.clustering.od_plots}

plot_growth_curves(dgobj, facet="cluster_assignment", color="genotype")

```

  
Of course one can play around with clustering by changing the UMAP or fPCA embedding parameters, and also the clustering parameters themselves.   
  
## Estimating growth parameters.  
  
Now that we have explored our data a bit, we can get some growth statistics from our data. In principle, we could do this directly after pre-processing. One of the main features of `DGrowthR` is htat it does not assume a specific functional form our growth curves. This means that we can flexibly model growth curves that do follow a typical sigmoidal shape. Briefly, a Gaussian Process (GP) is fit to our growth data (we use the [laGP](https://www.jstatsoft.org/article/view/v072i01) package for this), and the main growth statistics are gathered from the mean posterior. 
  
The growth parameters computed are:  
  
1. **max_growth**: The maximum growth achieved.  
2. **max_growth_time**: The Time at which **max_growth** was reached.  
3. **AUC**: The empirical area under the curve.  
4. **growth_loss**: The maximum decrease of growth after the moment of maximum growth.  
5. **max_growth_rate_timepoint**: The Time at which the maximum growth rate is acheived.  
6. **max_growth_rate**: The maximum specific growth rate.   
7. **doubling_time**: Equal to $log(2)/\text{max_growth_rate}$. 
8. **max_death_rate_timepoint**: The Time where the lowest growth rate is observed.  
9. **max_death_rate**: Maximum rate of death. Properly speaking it is the lowest growth rate.  
10. **lag_phase_timepoint**: The Time where lag phase ends and exponential growth begins. We find this as the maximum of the second derivative _before_ **max_growth_rate_timepoint**.
11. **growth_at_lag_phase**: Amount of growth when lag phase ends.  
12. **stationary_phase_timepoint**: The Time where the stationary phase starts. We find this as the minimum of the second derivative _after_ **max_growth_rate_timepoint**. In a typical growth curve, this would correspond to the beginning of the stationary phase. 
13 **growth_at_stationary_phase**: The amount of growth at **stationary_phase_timepoint**.  
  
**KEEP IN MIND** the values reported are in the units that were used during modelling. Since our data was log transformed during pre-processing, all growth parameters are in units of log(OD).  

  
To estimate growth parameters, we can use the `estimate_growth_parameters()` function.

```{r estimate.growth.params}

dgobj.curve_modelling <- estimate_growth_parameters(dgobj)

```



By default, the function models and gathers growth parameters for each individual curve and saves in the `@growth_parameters` slot. We can view it here.  
  
```{r view.curve_params}
dgobj.curve_modelling@growth_parameters %>% 
  rmarkdown::paged_table()
```
  
As you can see, the **gpfit_id** field corresponds to the **curve_id**. We actually update our low-dimensional embeddings with this data now.  


```{r update.and.view.curves}

# TODO: ADD THIS AS A FUNCTION?

# Gather growth parameters
growth_params <- dgobj.curve_modelling@growth_parameters %>% 
  rename("curve_id" = "gpfit_id")

# Add to metadata
dgobj.curve_modelling@metadata <- dgobj.curve_modelling@metadata %>% 
  left_join(growth_params, by="curve_id")

# Replot
plot_fpca(dgobj.curve_modelling, color = "AUC")

```
  
    
### Modelling groups of curves. 
  
One additional advantage of using GP to model our curves is that we can now model several growth curves at the same time! Instead of modelling one curve at a time, we can jointly model all replicate curves that were gathered in the same experimental conditions. For example, in [Brenzinger 2024](https://www.nature.com/articles/s41564-023-01556-y) they analyse the response of a $\Delta$CBASS strain against the wildtype when confronted with different compounds. We can jointly analyse all growth curves coming from the same genotype_compound combination.

```{r estimate.growth.by.cov}

dgobj.genotype_well <- estimate_growth_parameters(dgobj, 
                                                  model_covariate="genotype_well", # covariate in metadata to model
                                                  save_gp_data = TRUE, # Save the GP data
                                                  n_cores=3)
```
  
Now **gpfit_id** indicates each group modelled.
  
```{r}
dgobj.genotype_well@growth_parameters %>% 
  rmarkdown::paged_table()
```
    
    
Additionally, we also indicated that we should save the mean GP posterior. This can now be accessed in the `@gpfit_info` field. We can also now plot this information as below.


```{r}
plot_gp(dgobj.genotype_well, gpfit_id = "dCBASS_A11", show_input_od=TRUE)
```

  
Yet another great trick we can do is sample the posterior of the GP to gather an arbitrary number of synthetic growth curves. We can then gather the growth statistics from these sampled curves and get the mean and standard deviation of the growth parameters. 
  
Keep in mind that it only makes sense to do this when the modeled covariate contains more than one growth curve per group.

```{r estimate.growth.sampling, eval=FALSE}
dgobj.genotype_well.posterior <- estimate_growth_parameters(dgobj, 
                                                  model_covariate="genotype_well",
                                                  
                                                  predict_n_steps = 100, # Predict a lower number of GP points to speed up
                                                  
                                                  sample_posterior_gpfit=TRUE, # Sample the posterior to get mean and sd of each parameter
                                                  
                                                  sample_n_curves=10,  # Number of curves to sample from posterior. 10 for speed.
                                                  save_gp_data = FALSE, 
                                                  n_cores=4) # Increased number of cores
```
  
  
Now we can see that our growth statistics have a mean and standard deviation.

```{r views.posterior.growth params, eval=FALSE}
dgobj.genotype_well.posterior@growth_parameters %>% 
  rmarkdown::paged_table()
```
  
   
## Differential growth. 
  
One of the other main features of `DGrowthR` is the ability to compare growth curves gathered from different experimental conditions. This done via a likelihood ratio test. Briefly, two GP models are fit: a null model where all growth curves are modeled jointly, and an alternative model where an additional covariate is added to the design matrix indicating whether the growth curve come from the reference or treatment condition. Then the log-likelihood of these two models is compared, giving us the likelihood ratio test statistic.  
  
In `DgrowthR` all of these steps can be run with the `growth_comparison()` function. Here a comparison is indicated to the **comparison_info** parameter as a vector consisting of (metatada_field, treatment_value, reference_value). The metadata field should be a field in the metadata, and the treatment and reference values should be found in said field. These groups of growth curves are compared to eachother as described above.  
  
  
Following our example, we can compare how genotype affects the response to the same compound.

```{r gcomp}
object.comparison <- growth_comparison(object = dgobj, 
                                       
                                       # A comparison is indicated as metadata_field, treatment, reference.
                                       comparison_info = c("genotype_well", "dCBASS_O1" ,"wt_O1"),
                                       save_gp_data = TRUE)


# View result of comparison
growth_comparison_result(object.comparison) %>% 
  rmarkdown::paged_table()
```
  
  
Here we see the result of our comparison.   
  
1. **likelihood_ratio**: Indicates the test statistic for the likelihood ratio test. It is equal to **llik.alternative_model** -  **llik.null_model**.  
2. **llik.alternative_model**: The log-likelihood of the GP fit with the alternative model.  
3. **llik.null_model**: The log-likelihood of the GP fit with the null model.  
4. **AUC.treatment**: The AUC of the alternative model fit for the treatment growth curves.  
5. **AUC.reference**: The AUC of the alternative model fit for the reference growth curves. 
6. **AUC.FoldChange**: Equal to $\frac{\text{AUC.treatment}}{\text{AUC.reference}}$.  
7. **max_growth.treatment**: Maximum growth determined for the alternative model fit for the treatment growth curves.  
8. **max_growth.reference**: Maximum growth determined for the alternative model fit for the reference growth curves.  
9. **max_growth.FoldChange**: Equal to $\frac{\text{max_growth.treatment}}{\text{max_growth.reference}}$.  
10. **euclidean.distance**: The euclidean distance between the model fit for the treatment and alternative curves.  

  
We can get a better intuition by looking at a visual representation of our model. Below we see the fit of the null model, where all growth curves are modelled jointly, independent of the genotype. We can clearly see that the fit is poor. 


```{r plot.growth.comparison}

model_plots <- plot_growth_comparison(object.comparison)

model_plots$null
```

  
The alternative model fits the original growth curves much better, indicating that these two groups might be significantly different.

```{r growth.comparison.alt}
model_plots$alternative
```



Of course, up until this point we only have visual evidence for this significant difference. We can get some statistical significance by performing a permutation test. In `DGrowthR` the permutation test consists of randomly shuffling the label of each growth curve and computing the log-likelihood of the alternative GP model when these curves are shuffled. Doing this several times can give us an idea of the the distribution of the likelihood ratio test statistic when curves are randomly assigned. In this case we can set the **permutation_test** parameter and indicate the number of permutations we would like to perform. 
  

```{r perm.test.comparison}
object.comparison.permutation <- growth_comparison(object = dgobj, 
                                       comparison_info = c("genotype_well", "dCBASS_O1" ,"wt_O1"),
                                       permutation_test=TRUE,
                                       n_permutations = 100, # Here we use 100 for speed but should probably be more
                                       n_cores = 1,
                                       save_gp_data = FALSE)
```
  
As we can see below, a new field is added to the results **empirical_p.value**. This empirical p-value is calculated as the fraction of premuted test statistics that are larger than the observed test statistic. We can see below that the empirical_p.value is 0.009, which is quite good considering this is the lowest possible p-value in our setup.


```{r view.emp.result}
growth_comparison_result(object.comparison.permutation)
``` 
  
We can get a visual inspection of the null vs observed test statistic below

```{r emp.distribution}
plot_likelihood_ratio(object.comparison.permutation, num_bins = 50)
```

We can also plot the difference between the reference and treatment condition over time

```{r od.difference}
plot_od_difference(object.comparison)
```

## Multiple comparisons. 
  
Now let's imagine that we actually want to make several comparisons in one go. For that we can use the **multiple_comparisons** function. It is actually quite similar to the function above only that now we provide a list that contains all of the comparisons we want to make

```{r mult.comparisons}
contrast_list <- list(c("genotype_well", "dCBASS_O1" ,"wt_O1"),
                      c("genotype_well", "dCBASS_A1" ,"wt_A1"),
                      c("genotype_well", "dCBASS_B1" ,"wt_B1"))

# Returns a dataframe
multiple_df <- multiple_comparisons(dgobj,contrast_list, permutation_test = TRUE, n_permutations = 100)
```


Now we can view the results in one data.frame as shown below. There is also now an **pvalue.adjust** field which performs BH correction to the empirical_p.value field after all tests are done.


```{r mult.comparison.result}
multiple_df %>% 
  rmarkdown::paged_table()
```

That should be good to get you started!
